---
title: "MLS KeeperIQ — R Statistical Report"
subtitle: "Stability, rank disagreement, uncertainty, and New York Red Bulls trends"
author: "KeeperIQ analysis layer"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
library(tidyverse)
library(DBI)
library(duckdb)
library(boot)
library(scales)
library(gt)

source(file.path("R", "db.R"))
source(file.path("R", "analysis.R"))

# Keep the connection in the global knitr session — do not use on.exit()
# inside a setup chunk (it would disconnect when the chunk ends).
con <- connect_keeperiq()
```

## Purpose

This report is the R analysis companion to the Python/TypeScript KeeperIQ product.
It reads from the DuckDB relational database (`data/keeperiq.duckdb`) rather than
re-implementing the scoring model. The goals are to:

1. quantify year-to-year stability of adjusted goalkeeper impact
2. contrast KeeperIQ ranks with traditional goals-allowed ranks
3. bootstrap match-level uncertainty for selected keepers
4. inspect component correlations
5. profile New York Red Bulls goalkeepers specifically

## Data pulled through SQL

```{r load-data}
leaderboard <- DBI::dbGetQuery(con, "
  SELECT *
  FROM v_season_leaderboard
  ORDER BY season, rank_adjusted
") |> as_tibble()

components <- DBI::dbGetQuery(con, "
  SELECT *
  FROM v_component_matrix
") |> as_tibble()

disagreement <- DBI::dbGetQuery(con, "
  SELECT *
  FROM v_rank_disagreement
  WHERE season IN (2025, 2026)
") |> as_tibble()

match_form <- DBI::dbGetQuery(con, "
  SELECT *
  FROM v_match_form
  WHERE season IN (2025, 2026)
") |> as_tibble()

talent <- DBI::dbGetQuery(con, "
  SELECT *
  FROM v_talent_board
  ORDER BY talent_rank
") |> as_tibble()

red_bulls <- leaderboard |>
  filter(str_detect(tolower(coalesce(team_name, "")), "red bull") |
           team_abbreviation %in% c("NYRB", "RBNY"))

knitr::kable(
  leaderboard |>
    count(season, sample_status) |>
    pivot_wider(names_from = sample_status, values_from = n, values_fill = 0),
  caption = "Goalkeeper counts by season and sample status"
)
```

## 1. Year-to-year stability

How stable is reliability-adjusted total G+/96 for keepers who appear in
adjacent seasons with provisional-or-better minutes?

```{r yoy}
stability <- yoy_stability(leaderboard)
stability |>
  gt() |>
  fmt_number(columns = spearman, decimals = 3) |>
  tab_header(
    title = "Spearman stability of adjusted total G+/96",
    subtitle = "Qualified and provisional goalkeepers only"
  )
```

```{r yoy-plot}
shared <- leaderboard |>
  filter(sample_status %in% c("qualified", "provisional"), season %in% c(2025, 2026)) |>
  select(player_id, player_name, season, adjusted_total_p96) |>
  pivot_wider(names_from = season, values_from = adjusted_total_p96, names_prefix = "y") |>
  drop_na()

ggplot(shared, aes(x = y2025, y = y2026)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.8, colour = "#0B3D2E", size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "#C4A35A", fill = "#C4A35A33") +
  labs(
    title = "2025 vs 2026 adjusted total G+/96",
    x = "2025 adjusted G+/96",
    y = "2026 adjusted G+/96",
    caption = "Dashed line is y = x. Points are keepers with provisional+ minutes in both seasons."
  ) +
  theme_minimal(base_size = 12)
```

## 2. Goals-allowed rank versus KeeperIQ rank

A disagreement identifies goalkeepers whose context-adjusted evaluation differs
from the traditional result. It does **not** automatically prove one metric is
correct.

```{r disagreement-plot}
ggplot(
  disagreement |> filter(season == 2025, !is.na(rank_disagreement)),
  aes(x = goals_allowed_rank, y = keeperiq_rank, colour = rank_disagreement)
) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_colour_gradient2(low = "#B33A3A", mid = "grey70", high = "#1F7A4C") +
  scale_x_reverse() +
  scale_y_reverse() +
  labs(
    title = "2025: goals-allowed rank vs KeeperIQ rank",
    x = "Goals allowed rank (lower = fewer goals conceded / 96)",
    y = "KeeperIQ rank (lower = better)",
    colour = "Rank gap\n(GA rank − IQ rank)"
  ) +
  theme_minimal(base_size = 12)
```

```{r disagreement-table}
disagreement |>
  filter(season == 2025) |>
  mutate(abs_gap = abs(rank_disagreement)) |>
  slice_max(abs_gap, n = 12, with_ties = FALSE) |>
  select(
    player_name, team_abbreviation, keeperiq_rank, goals_allowed_rank,
    rank_disagreement, keeperiq, adjusted_total_p96, goals_conceded_p96, minutes
  ) |>
  gt() |>
  fmt_number(columns = c(keeperiq, adjusted_total_p96, goals_conceded_p96), decimals = 2) |>
  tab_header(title = "Largest 2025 rank disagreements")
```

## 3. Bootstrap uncertainty for selected keepers

Match-level nonparametric bootstrap of mean match G+/96 for high-minute 2026
keepers. This is an R-side uncertainty check complementary to the Python
pipeline's own bootstrap intervals.

```{r bootstrap}
targets <- leaderboard |>
  filter(season == 2026, sample_status == "qualified") |>
  slice_min(rank_adjusted, n = 5, with_ties = FALSE)

boot_rows <- purrr::map_dfr(targets$player_name, function(name) {
  rates <- match_form |>
    filter(player_name == name, season == 2026) |>
    pull(ga_total_p96) |>
    purrr::discard(is.na)
  fit <- bootstrap_player_mean(rates, R = 1000L)
  tibble(
    player_name = name,
    n_matches = length(rates),
    mean_match_ga_p96 = fit$t0,
    low_90 = fit$conf[1],
    high_90 = fit$conf[2]
  )
})

boot_rows |>
  gt() |>
  fmt_number(columns = c(mean_match_ga_p96, low_90, high_90), decimals = 3) |>
  tab_header(
    title = "Bootstrap 90% intervals for mean match G+/96",
    subtitle = "Top five qualified 2026 goalkeepers by KeeperIQ rank"
  )
```

## 4. Component correlations (2025)

```{r component-cors}
cmat <- component_cors(components, 2025)
cmat |>
  as.data.frame() |>
  rownames_to_column("component") |>
  gt() |>
  fmt_number(decimals = 2) |>
  tab_header(title = "Spearman correlations among adjusted component rates — 2025")
```

```{r component-heatmap}
cmat_df <- as.data.frame(as.table(cmat)) |>
  as_tibble() |>
  rename(component_a = Var1, component_b = Var2, rho = Freq)

ggplot(cmat_df, aes(component_a, component_b, fill = rho)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(limits = c(-1, 1), midpoint = 0) +
  coord_fixed() +
  labs(title = "2025 component correlation heatmap", x = NULL, y = NULL, fill = "ρ") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5. New York Red Bulls goalkeeper trends

```{r rbn-table}
red_bulls |>
  arrange(season, rank_adjusted) |>
  select(
    season, player_name, sample_status, minutes, keeperiq,
    adjusted_total_p96, observed_total_p96, goals_conceded_p96,
    save_pct, goals_prevented_p96, rank_adjusted, rank_disagreement, archetype_label
  ) |>
  gt() |>
  fmt_number(
    columns = c(keeperiq, adjusted_total_p96, observed_total_p96,
                goals_conceded_p96, save_pct, goals_prevented_p96),
    decimals = 2
  ) |>
  tab_header(title = "New York Red Bulls goalkeepers in KeeperIQ")
```

```{r rbn-form}
rbn_names <- unique(red_bulls$player_name)
rbn_form <- match_form |>
  filter(player_name %in% rbn_names, season == 2026) |>
  arrange(date_utc)

if (nrow(rbn_form) > 0) {
  ggplot(rbn_form, aes(x = as.Date(date_utc), y = rolling_5_match_ga_p96, colour = player_name)) +
    geom_hline(yintercept = 0, colour = "grey70") +
    geom_line(linewidth = 1) +
    geom_point(aes(y = ga_total_p96), alpha = 0.35) +
    labs(
      title = "Red Bulls rolling 5-match G+/96 — 2026",
      x = "Match date",
      y = "Rolling G+/96",
      colour = "Goalkeeper",
      caption = "Points are individual match rates; lines are 5-match rolling averages."
    ) +
    theme_minimal(base_size = 12)
} else {
  cat("No 2026 Red Bulls match-form rows available in the current database extract.")
}
```

## 6. Current-talent weights for Red Bulls keepers

```{r rbn-talent}
talent |>
  filter(player_name %in% rbn_names) |>
  select(
    talent_rank, player_name, keeperiq_talent, talent_total_p96,
    weight_league_prior, weight_prior_season, weight_live_season,
    prior_season_minutes, live_season_minutes, prior_source
  ) |>
  gt() |>
  fmt_number(columns = c(keeperiq_talent, talent_total_p96), decimals = 2) |>
  fmt_percent(
    columns = c(weight_league_prior, weight_prior_season, weight_live_season),
    decimals = 1
  ) |>
  tab_header(title = "Current-talent blend — Red Bulls goalkeepers")
```

## Limitations

- Goals Added is itself a model estimate from American Soccer Analysis.
- R bootstraps here use match-level mean G+/96 and are complementary to, not
  identical with, the Python pipeline's shrinkage-aware bootstrap.
- Roster affiliation in the leaderboard view reflects ASA season team ids;
  mid-season transfers can complicate club labels.
- Small samples remain noisy even after reliability adjustment.

## Session info

```{r session}
sessionInfo()
```

```{r cleanup, include=FALSE}
if (exists("con") && DBI::dbIsValid(con)) {
  DBI::dbDisconnect(con, shutdown = TRUE)
}
```
